A Hexagon Tile Map Algorithm for Displaying Spatial Data

### 2023-07-13 kobakian-cook.tex

% !TeX root = RJwrapper.tex
\title{A Hexagon Tile Map Algorithm for Displaying Spatial Data}
\author{by Stephanie Kobakian, Dianne Cook, and Earl Duncan}

@@

Over time, the density of living in major cities has increased as residents have gathered to live near urban areas (Dorling 2011). The division of these residents into approximately equal population areas results in dramatically different square meterage of administrative geographic units, such as states or electorates. In a choropleth map display, the geographic units are coloured to represent the value of the statistic for each unit (Tufte 1990). This can cause some values to be emphasised over others, and allows choropleth map displays to misrepresent the spatial distributions of human related statistics due to area-size bias (Skowronnek 2016).

The Australian Cancer Atlas is an online, interactive display of Australian cancer statistics in each of the Statistical Areas at Level 2 (SA2s), used by the Australian Bureau of Statistics (2011). The estimated standardised incidence ratios (SIRs) of thyroid cancer
for females was downloaded from the publicly accessible Australian Cancer Atlas website and presented in Figure \ref{fig:choro}, a choropleth map that uses colour to display the distribution. The Australian choropleth map display draws attention to the expanse of dark and light blue areas across the rural communities in all states. The SA2s on the east coast around Brisbane and in northern New South Wales stand out as more orange and red. However, this display neglects the vast amount of Australia residents living in the densely populated capital cities.

@@

The solutions to this visualisation problem begin with the geography. Alternative maps can be created to shift the focus from land area and shape to the value of the statistics (Dougenik, Chrisman, and Niemeyer 1985). Each style of cartogram applies a different transformation to the geographic areas, to highlight the values of the statistic of interest. Alternative maps can result in a distortion of the map space to represent features of the distribution across the areas (Dougenik, Chrisman, and Niemeyer 1985) as the statistic of interest is used to determine the cartogram layout.

Alternative mapping methods, like cartograms implemented in \CRANpkg{cartogram}, allow increased understanding of the spatial distribution of a variable across the population, by fairly representing the population in each administrative area (Levison and Haddon Jr 1965). This acknowledges that the number of residents can be different and also recognises that each area, or person within it is equally important.

@@

The familiar choropleth map display is often used to allow the user to orient themselves. However, the emphasis on the land mass diminishes the features of the distribution in densely populated communities due to the small size on the display (Dorling 2011). The unique shapes of boundaries can be helpful for orienting users but may not contribute to their understanding of the spatial disease distribution as many of the communities are not visible in a choropleth display (Dorling 2012).

@@

There are several established alternative visualisation methods. Rectangular cartograms (Kreveld and Speckmann 2007) and Dorling cartograms (Dorling 2011) implemented in \pkg{cartogram} and tile maps implemented in \CRANpkg{tilemaps}, all use one simple shape to represent each geographic unit. They all minimise the emphasis on the size or shape of the geographic areas. These alternative map displays focus on the relationship between neighbours, attempting to preserve connections, and disregard the unique shapes of the administrative boundaries. Figure \ref{fig:tasdisplays} shows a collection of alternative map displays, this includes a) a contiguous cartogram, b) a non-contiguous cartogram, and c) a Dorling cartogram.

@@

When communicating information that is relevant to the population, each member of the population can be given equal representation by transforming the map (Dorling 2012). The connected communities can be recognised by allowing the boundaries to maintain connection in the transformed display. The contiguous cartogram displayed in Figure \ref{fig:tasdisplays}a draws attention to smaller geographic units when they are rescaled according to the population (Walter 2001). These new shapes can now be coloured to represent a second variable. The algorithm uses the geographic shape of the areas and iterates toward sizing the areas to represent the population. This display can create twisted and unfamiliar shapes from the geographic units as the algorithms must satisfy the topology conditions, especially when there are communities located geographically far from their neighbours (Dorling 2012).

The non-contiguous cartogram in Figure \ref{fig:tasdisplays}b also uses the population to rescale the geographic units. Unlike the contiguous cartogram, the SA2 areas maintain their geographic shape, but they may not retain the connection with their neighbours. The population of the SA2 areas is used to scale the geographic units in the non-contiguous cartogram (Olson 1976). The amount of background space can be meaningful in non-contiguous cartograms (Keim et al. 2002). The size disparity between rural areas and urban areas result in reasonable display within the south eastern city of Hobart, but these units are not visible in the context of the Tasmania state map. Depending on the difference between the population and geographic land size, the amount of white space can also prevent meaningful understanding of the distribution (Dorling 2012).

The Dorling cartogram presents each geographic unit as a circle, the size of the circle is scaled according to the population value of each area (Dorling 2011). Figure \ref{fig:tasdisplays}c shows the Tasmania SA2 areas as an individual circle located as close as possible to the geographic centroid location. This map draws attention to the collection of coastal cities in Tasmania that were not apparent in Figure \ref{fig:tasdisplays} a or b. This display also highlights that the approximation of equal population is not easy to fulfil, as there is some disparity in the sizes of the circles.

It is also common to treat all geographic units as equal, so that regardless of the population or other information, to generate equal size spatial polygons. The Dorling algorithm can be induced to do this, to make equallly sized circles. Cano et al. (2015) describes a group of alternative maps called ``mosaic cartograms'' for tile displays. The geography of the USA has been transformed into a square-binned tile map manually and made available in the R package \CRANpkg{statebins} (Rudis 2020).The algorithm described below is an automatic layout of hexagon tiles, where a continguity constraint is relaxed. This gives equal weight to each geographic area even if the population differs.

@@

Hexagon tile maps can be useful to understand a distribution across a collection of geographic areas. However, these maps are not easy to create manually, especially as the number of areas increases. This algorithm was created to automate this process, and reduce the manual load involved in creating and implementing alternative displays. This allows map makers and data communicators to spend their time choosing the most effective display.

The \pkg{sugarbag} package contains a set of functions that help R users to create a hexagon tile map. The algorithm presented in the \pkg{sugarbag} package operates on a set of simple feature geometry objects , known as \CRANpkg{sf} objects(Pebesma 2018). This package allows R users to create \pkg{sf} objects by importing polygons stored in various formats. Users should provide a set of polygons that define geographic units by their administrative boundaries. The functions arrange the geographic units in order of proximity to a set of locations provided, such as the centre of major cities. The centroid location of each geographic unit is used to measure the proximity. It emphasises the major cities as population hubs, rather than emphasizing the size of large, rural geographic units.

A user can tweak the parameters of the hexagon map using the additional arguments to the \code{create\_hexmap} function, but these options may affect the speed of the algorithm.
The hexagon size may need adjustments depending on the density of the population; users can provide an appropriate hexagon size and re-run the algorithm. The buffer distance may need to be increased if the coastal cities need to extend beyond the geographic land mass.

@@

If a user would like to perform each of the steps of the algorithm themselves, the necessary inputs will change for each function. For example, the \code{allocate} function can be used directly when a user wishes to use a set of centroids, rather than polygons.

The set of SA2 polygons for Tasmania, using the 2011 definition of SA2s, is accessed from the \pkg{absmapsdata} package. A single column of the data set is used to identify the unique areas. In this case, the unique SA2 names for each SA2 have been used.

@@

@@

@@

@@

\address{%
Stephanie Kobakian\\
\\%
Sydney, Australia\\
%
%
%
\href{mailto:stephanie.kobakian@gmail.com}{\nolinkurl{stephanie.kobakian@gmail.com}}%
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
%
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}%
}

\address{%
Earl Duncan\\
Queensland University of Technology\\%
School of Mathematical Sciences\\
%
%
%
\href{mailto:earl.duncan@qut.edu.au}{\nolinkurl{earl.duncan@qut.edu.au}}%
}

>>>>

% !TeX root = RJwrapper.tex
\title{A Hexagon Tile Map Algorithm for Displaying Spatial Data}

\author{by Stephanie Kobakian, Dianne Cook, and Earl Duncan}

@@

Over time, the population density in major cities has increased as residents have gathered to live near urban areas (Dorling 2011). Populations tend to be distributed unevenly across geographic regions. When comparing sets of administrative geographic units such as states or electorates, area size is seldom proportional to the population size. In a choropleth map display, the geographic units are coloured to represent the value of the statistic for each unit (Tufte 1990). This can cause some values to be emphasised over others, and allows choropleth map displays to misrepresent the spatial distributions of human related statistics due to area-size bias (Skowronnek 2016).

The Australian Cancer Atlas is an online, interactive display of Australian cancer statistics in each of the Statistical Areas at Level 2 (SA2s), used by the Australian Bureau of Statistics (2011). The dataset of estimated standardised incidence ratios (SIRs) of thyroid cancer for females was downloaded from the publicly accessible Australian Cancer Atlas website and presented in Figure \ref{fig:choro}, a choropleth map that uses colour to display the distribution. The Australian choropleth map display draws attention to the expanse of dark and light blue areas across the rural communities in all states. The SA2s on the east coast around Brisbane and in northern New South Wales stand out as more orange and red. However, this display neglects the vast amount of Australian residents living in the densely populated capital cities.

@@

The solutions to this visualisation problem begin with the geography. Alternative maps can be created to shift the focus from land area and shape to the value of the statistics (Dougenik, Chrisman, and Niemeyer 1985) in a cartogram display. Cartogram methods apply different transformations to the geographic areas, to highlight the values of the statistic of interest. Alternative maps can result in a distortion of the map space to represent features of the distribution across the areas (Dougenik, Chrisman, and Niemeyer 1985) as the statistic of interest is used to determine the cartogram layout.

Alternative mapping methods, like cartograms implemented in \CRANpkg{cartogram} (Jeworutzki 2020) for R (R Core Team 2012), promote better understanding of the spatial distribution of a variable across the population, by representing the population in each administrative area fairly (Levison and Haddon Jr 1965). This acknowledges that the number of residents can be different and also recognises that each area, or person within it is equally important.

@@

Typically, chloropleth maps are used to orient the users within the geographic context. However, the strong salience of the land mass can hide or downplay the features of the distribution in densely populated communities due to the small size on the display (Dorling 2011). The unique shapes of boundaries can be helpful for orienting users but may not contribute to their understanding of the spatial disease distribution as many of the communities are not visible in a choropleth display (Dorling 2012).

@@

There are several established alternative visualisation methods. Rectangular cartograms (Kreveld and Speckmann 2007) and Dorling cartograms (Dorling 2011) implemented in \pkg{cartogram} and tile maps implemented in \CRANpkg{tilemaps} (Rosenberg 2020), all use one simple shape to represent each geographic unit. They all minimise the salience of the size or shape of geographic areas. These alternative map displays highlight the relationship between neighbours, preserve connections, and discard the unique shapes of the administrative boundaries. Figure \ref{fig:tasdisplays} shows a collection of alternative map displays, this includes a) a contiguous cartogram, b) a non-contiguous cartogram, and c) a Dorling cartogram.

@@

When communicating information that is relevant to the population, each member of the population can be given equal representation by transforming the map (Dorling 2012). The connectedness of the units can be preserved by using transformations that maintain the connection between boundaries. The contiguous cartogram displayed in Figure \ref{fig:tasdisplays}a draws attention to smaller geographic units when they are rescaled according to the population (Walter 2001). These new shapes can now be coloured to represent a second variable. The algorithm uses the geographic shape of the areas and iterates toward sizing the areas to represent the population. This display can create twisted and unfamiliar shapes from the geographic units as the algorithms must satisfy the topology conditions, especially when there are communities located geographically far from their neighbours (Dorling 2012).

The non-contiguous cartogram in Figure \ref{fig:tasdisplays}b also uses the population to rescale the geographic units. Unlike the contiguous cartogram, the SA2 areas maintain their geographic shape, but they may not retain the connection with their neighbours. The population of the SA2 areas is used to scale the geographic units in the non-contiguous cartogram (Olson 1976). The amount of background space can be meaningful in non-contiguous cartograms (Keim et al. 2002). However, the relative size of units in comparison to the reference unit can lead to scaling issues. The size disparity between rural areas and urban areas result in reasonable display within the south eastern city of Hobart, but these units are not visible in the context of the Tasmania state map. Depending on the difference between the population and geographic land size, the amount of white space can also prevent meaningful understanding of the distribution (Dorling 2012).

The Dorling cartogram presents each geographic unit as a circle, the size of the circle is scaled according to the population value of each area (Dorling 2011). Figure \ref{fig:tasdisplays}c shows the Tasmania SA2 areas as an individual circle located as close as possible to the geographic centroid location. This map draws attention to the collection of coastal cities in Tasmania that were not apparent in Figure \ref{fig:tasdisplays} a or b. This display also highlights the difference in the population of each unit, as there is some disparity in the sizes of the circles.

Another common practice is to use equal-sized polygons and treat all geographic units as equal. The Dorling algorithm can be induced to do this, to make equallly sized circles. Cano et al. (2015) describes a group of alternative maps called ``mosaic cartograms'' for tile displays. Similarly, the geography of USA has been transformed into a square-binned tile map manually and made available in the R package \CRANpkg{statebins} (Rudis 2020).The algorithm described below generates an automatic layout of hexagon tiles, where a continguity constraint is relaxed. This gives equal weight to each geographic area even if the population differs.

@@

Hexagon tile maps can be useful for visualising distributions across a collection of geographic areas. However, these maps are not easy to create manually, especially as the number of areas increases. This algorithm was developed to automate the process, and reduce the workload involved in creating and implementing alternative displays. This allows map makers and data communicators to spend their time choosing the most effective display.

The \pkg{sugarbag} package contains a set of functions that help R users to create a hexagon tile map. The algorithm presented in the \pkg{sugarbag} package operates on a set of simple feature geometry objects , known as \CRANpkg{sf} objects (Pebesma 2018). This package allows R users to create \pkg{sf} objects by importing polygons stored in various formats. Users should provide a set of polygons that define geographic units by their administrative boundaries. The functions arrange the geographic units in order of proximity to a set of locations provided, such as the centre of major cities. The centroid location of each geographic unit is used to measure the proximity. It emphasises the major cities as population hubs, rather than emphasizing the size of large, rural geographic units.

The user can tweak the parameters of the hexagon map using additional arguments to the \code{create\_hexmap} function, but these options may affect the speed of the algorithm.
The hexagon size may need adjustments depending on the density of the population; users can provide an appropriate hexagon size and re-run the algorithm. The buffer distance may need to be increased if the coastal cities need to extend beyond the geographic land mass.

@@

If a user would like to perform steps of the algorithm themselves, additional user input will be needed for the functions that perform each step. For example, if the user wishes to use a set of centroids, rather than polygons, the \code{allocate} function can be used directly.

The set of SA2 polygons for Tasmania, using the 2011 definition of SA2s, was accessed from the \pkg{absmapsdata} package (Mackey 2022). A single column of the data set is used to identify the unique areas. In this case, the unique SA2 names for each SA2 have been used.

@@

\leavevmode\vadjust pre{\hypertarget{ref-cartogram}{}}%
Jeworutzki, Sebastian. 2020. \emph{Cartogram: Create Cartograms with r}. \url{https://CRAN.R-project.org/package=cartogram}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-absmapsdata}{}}%
Mackey, Will. 2022. \emph{Absmapsdata: A Catalogue of Ready-to-Use ASGS (and Other) Sf Objects}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
R Core Team. 2012. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{http://www.R-project.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-tilemaps}{}}%
Rosenberg, Kaelyn. 2020. \emph{Tilemaps: Generate Tile Maps}. \url{https://CRAN.R-project.org/package=cartogram}.

@@

\address{%
Stephanie Kobakian\\
\\%
Sydney, Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-4616-0048}{0000-0002-4616-0048}}\\%
\href{mailto:stephanie.kobakian@gmail.com}{\nolinkurl{stephanie.kobakian@gmail.com}}%
}

\address{%
Dianne Cook\\
Monash University\\%
Department of Econometrics and Business Statistics\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-3813-7155}{0000-0002-3813-7155}}\\%
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}%
}

\address{%
Earl Duncan\\
Queensland University of Technology\\%
School of Mathematical Sciences\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-5146-7810}{0000-0002-5146-7810}}\\%
\href{mailto:earl.duncan@qut.edu.au}{\nolinkurl{earl.duncan@qut.edu.au}}%
}
